---
title: "worlddatr" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{worlddatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  fig.align = "center",
  comment = "#>"
)
```

`worlddatr` makes creating choropleth maps easier and quicker so you can spend more time of your research instead of data cleaning. Often, you will want to analyse global trends in data, include metrics such as income level of the country and plot all of this information on a map of the world. However, these various data sources use different variables, country names and codes, consequently, the data is time-consuming and burdensome to use, especially if you are new to programming langauages such as R. For instance:

-   Various spellings of a country/territory (i.e. North Korea, N.Korea, Democratic People's Republic of Korea)
-   English or native translations, non-ASCII Characters (i.e. Spain, España)
-   Different definitions or groupings of countries/territory (i.e. Channel Islands grouped together, or Jersey, Guernsey etc. listed separately)

Typically, each researcher wanting to plot the globe would need to go through this data cleaning process, we've done this so you don't have to and packaged it into an R package, `worlddatr`. Key features include:

-   Pre-processed Datasets: Access to pre-cleaned datasets that are ready to merge with your data, reducing the time spent on data preparation.
-   One-command Utilty: Functions that allow you to produce maps with a single command, while still providing control over customisation options such as colours, titles, and legends.
-   Open Source and Community-driven: Being open-source, worlddatr invites users to build upon its code, tailor visualisations to their specific needs, and contribute improvements back to IDDO, enhancing the tool for the entire research community.

## Installing worlddatr

You can install the `worlddatr` package from [GitHub](https://github.com/Infectious-Diseases-Data-Observatory/worlddatr), there you can find the code required.

```{r setup}
library(worlddatr)
```

## Datasets

There are two datasets in the package, `world_income`, which has a row per country/territory and provides the country ISO codes, name and income group. This has 250 rows and 7 columns. `alpha_3_code`, `alpha_2_code` and `numeric` refer to the ISO standardised country codes, `country` and `economy` list the names of the country from ISO and World Bank datasets respectively, while `income_group` provides the income infomation for each entity. `redcap_number` is also included as a classification number for IDDO REDCap databases, which is an alphabetical, English list of countries, note this is not standardised so check that you data matches this list before using the `redcap_number`.

```{r}
head(world_income)
```

The second dataset merges the previous `world_income` with the `ggplot2` `map_data()` function, with cleaning steps applied to the latter, resulting in a 99,338 row, 13 column dataset which, in addition, contains the longitude and latitude of the borders of the countries and territories which we desire to plot. `region`, `subregion`, `group` and `order` are used to identify the country borders.

```{r}
head(world_map)
```

These datasets are also stored as csv files in the GitHub for use.

## Functions

As well as the data, we introduce some functions which can assist making the maps, so that regular users and beginners alike can utilise this tool. We'll show `create_participant_map()` which allows the user to make a choropleth for the number of participants per country in a dataset, there is also `create_studies_map()` which is very similar but the default legend refers to studies as opposed to particiapnts, the rest of the parameters work the same way.

Your dataset will need to include what country the participant or study (or other entity) is from, in the 3 digit ISO code format. The function will group the data by the country and then summarise the data in a `ggplot` using the `geom_polygon()` function. If your data does not have the 3 digit ISO code, `left_join()` your dataset with `world_income` with the key being the country code or name you do have, then you'll have the `alpha_3_code` in your data and use this as the `country_col` below. 

#### Create Synthetic Data for Example

```{r}
set.seed(123) # for reproducibility

countries <- sample(world_income$alpha_3_code, 100, replace = FALSE)

probabilities <- runif(length(countries))
probabilities <- probabilities / sum(probabilities)

country_data <- data.frame(COUNTRY = sample(countries, 10000, replace = TRUE, prob = probabilities))
```

#### Using the Functions

Now we can take our data, `country_data`, indicate the name of the column where the three digit ISO country code exists (`country_col`) and run the command.

```{r}
create_participant_map(data        = country_data,
                       country_col = "COUNTRY")
```

This visualises the example data, along with legend and the total number, n, of participants. Although, perhaps you do not want n to be displayed and you'll note that Antarctica is not present, using extra parameters we can change that.

```{r}
create_participant_map(data        = country_data,
                       country_col = "COUNTRY",
                       include_n   = FALSE,
                       include_ATA = TRUE)
```

The title and subtitle can be added to the plot to provide more information about your graphic, and the legend title can be changed, or removed by leaving the option blank.

```{r}
create_participant_map(data        = country_data,
                       country_col = "COUNTRY",
                       title       = "Number of Participants in Studies across Countries",
                       subtitle    = "Data from studies registered between 2000 - 2024",
                       legend      = "Some Legend")
```

Blue not your colour? No problem, customise the colours of the countries, borders, text and background easily.

```{r}
create_participant_map(data        = country_data,
                       country_col = "COUNTRY",
                       colour_high = "#FF3C38",
                       colour_low  = "#FFCFCF",
                       colour_default = "#F0DFEE",
                       colour_borders = "#20A39E",
                       colour_background = "#DAEDBD",
                       colour_text = "purple")
```

You can change the scale breaks on the legend using `scale_breaks` and log transform the scale using `log_scale`.

```{r}
create_participant_map(data        = country_data,
                       country_col = "COUNTRY",
                       scale_breaks = c(10, 50, 100, 180, 200))

create_participant_map(data        = country_data,
                       country_col = "COUNTRY",
                       log_scale   = TRUE,
                       legend      = "Log Scale of Participants")
```

This does not intend to be the only solution and we encourage you to use the code in the GitHub to develop graphs more specific to your needs, this hopes to provide some inspiration, without spending hours to see if it works. If you have suggestions, issues or want to contribute to the package contact us or see more on the GitHub page.
